---
title: "Reproducing Karin et al's longitudal model of senescent cell growth"
output:
  pdf_document: default
  html_document:
    df_print: paged
params:
  data_dir: ../dat/
  refit_models: F
---

```{r include=FALSE}
library(bayesplot)
library(brms)
library(dplyr)
library(foreach)
library(ggplot2)
library(ggpubr)
library(latex2exp)
library(loo)
library(purrr)
library(reshape2)
library(scales)
library(stringr)
library(tidyverse)

# Stan-specific stuff
library(bridgesampling)
library(rstan)
library(gridExtra)
rstan_options (auto_write = TRUE)
options(mc.cores = parallel::detectCores())
set.seed(3) # for reproductibility
```

# Longitudinal Data
## Loading and visualizing the data
```{r}
snc_counts_fname <- str_c(
  params$data_dir, "raw", "senescent_cells_and_the_dynamics_of_aging__fig_2.csv",
  sep="/"
)
raw_snc_counts <- read.csv(snc_counts_fname, header=TRUE) %>%
  drop_na()
snc_counts <- melt(
  data = raw_snc_counts, 
  id.vars = "mouse.week", 
  measure.vars = c(
    "X8", "X16", "X24", "X32", "X40", "X48", "X56", "X64", "X72", "X80"
  ),
  value.name = "TBL_x_10_5"
)
snc_counts <- snc_counts %>%
  rename(week = "variable", mouse = "mouse.week")
snc_counts$week <- map_int(snc_counts$week, function (week) { return(as.integer(substr(week, 2, str_length(week)))) })
length(unique(snc_counts$week))
```

```{r}
ggplot(
  data = snc_counts, 
  aes(x = week, y = TBL_x_10_5, color = mouse, group = mouse)
) +
  xlab("Week") + ylab(TeX("$ TBL \\times 10^5 $")) +
  scale_color_gradientn(colours = rainbow(5)) + geom_line() + geom_point()
max(snc_counts$mouse)
```


```{r}
norm_const <- mean(
  snc_counts[snc_counts$week %in% c(8, 16, 24),]$TBL_x_10_5, na.rm = T
)
snc_counts$norm_abundance <- snc_counts$TBL_x_10_5 / norm_const
mean_abundances <- aggregate(
  snc_counts$norm_abundance, list(snc_counts$week), mean, na.rm = T
)$x
stddev_abundances <- aggregate(
  snc_counts$norm_abundance, list(snc_counts$week), sd, na.rm = T
)$x
mean_stddev_abundances <- data.frame(
  mean = mean_abundances,
  stddev = stddev_abundances,
  week = unique(snc_counts$week)
)
p <- ggplot(mean_stddev_abundances, aes(x=week, y=mean)) + geom_point() +
    geom_errorbar(aes(ymin=mean-stddev/2, ymax=mean+stddev/2), width=.2,
                 position=position_dodge(0.05))
print(p)
```

```{r}
mean_abundances[2]
```


# Fitting our three models on longitudinal data
## Data processing
```{r}
# Time series of SnC counts
mice <- unique(snc_counts$mouse)
weeks <- unique(snc_counts$week)
n_mice <- length(mice)
n_weeks <- length(weeks)
snc_abundance <- matrix(nrow = n_mice, ncol=n_weeks)

for (m in 1:n_mice) {
  for (t in 1:n_weeks) {
    mouse <- mice[m]
    week <- weeks[t]
    mouse_week_index <- which(snc_counts$mouse == mouse & snc_counts$week == week)[1]
    snc_abundance[m, t] <- snc_counts$norm_abundance[mouse_week_index]
  }
}

# times
t <- sort(weeks) # 8 hard-coded to match data
t0 = 0
t

# initial conditions
y0 <- c(1)
dim(y0) <- (1)
y0_oos_young <- mean_abundances[1] + 50;
dim(y0_oos_young) <- 1

# data for Stan
data_snc <- list(
  n_weeks = n_weeks, 
  n_mice = n_mice, 
  y0 = y0, 
  t0 = t0, 
  ts = t, 
  cells = snc_abundance,
  n_weeks_oos_young = 5.,
  y0_oos_young = y0_oos_young,
  t0_oos_young = 8.
)

saveRDS(snc_counts, "saved_fit/snc_counts.RDS")
saveRDS(data_snc, "saved_fit/data_snc.RDS")

# number of MCMC steps
niter <- 2000
```

## MCMC sampling
### SR model

```{r}
snc_sr_model <- stan_model("snc_sr_model.stan")
if (params$refit_models) {
  fit_snc_sr_model <- sampling(
    snc_sr_model,
    data = data_snc,
    iter = niter,
    chains = 4,
    control = list(adapt_delta = .99)
  )
  saveRDS(fit_snc_sr_model, "saved_fit/fit_snc_sr_model.RDS")
} else {
  fit_snc_sr_model <- readRDS("saved_fit/fit_snc_sr_model.RDS")
}
```



### USR model

```{r}
if (params$refit_models) {
  model <- stan_model("snc_usr_model.stan")
  fit_snc_usr_model <- sampling(
    model,
    data = data_snc,
    iter = niter,
    chains = 4,
    control = list(adapt_delta = .99)
  )
  saveRDS(fit_snc_usr_model, "saved_fit/fit_snc_usr_model.RDS")
} else {
  fit_snc_usr_model <- readRDS("saved_fit/fit_snc_usr_model.RDS")
}
```

### SIS model
```{r}
if (params$refit_models) {
  model <- stan_model("snc_sis_model.stan")
  fit_snc_sis_model <- sampling(
    model,
    data = data_snc,
    iter = niter,
    chains = 4,
    control = list(adapt_delta = .99)
  )
  saveRDS(fit_snc_sis_model, "saved_fit/snc_sis_model.RDS")
} else {
  fit_snc_sis_model <- readRDS("saved_fit/snc_sis_model.RDS")
}
```

## Fit diagnostics
```{r}
sr_pars=c('eta', 'beta', 'sigma', 'lp__', 'log_likelihood')
print(fit_snc_sr_model, pars = sr_pars, digits_summary = 4)
```


```{r}
usr_pars=c('eta', 'beta', 'sigma', 'log_likelihood', 'lp__')
print(fit_snc_usr_model, pars = usr_pars, digits_summary = 4)
```

```{r}
sis_pars=c('eta', 'beta', 'sigma', 'alpha', 'll', 'log_likelihood', 'lp__')
print(fit_snc_sis_model, pars = sis_pars, digits_summary = 4)
```


```{r}
pairs(fit_snc_sr_model, pars=sr_pars)
```


```{r}
pairs(fit_snc_usr_model, pars = usr_pars)
```

```{r}
pairs(fit_snc_sis_model, pars = sis_pars)
```


```{r}
stan_dens(fit_snc_sr_model, pars = sr_pars, separate_chains = TRUE)
```

```{r}
stan_dens(fit_snc_usr_model, pars = usr_pars, separate_chains = TRUE)
```

```{r}
stan_dens(fit_snc_sis_model, pars = sis_pars, separate_chains = TRUE)
```

## Trajectory plots
```{r}
snc_sr_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sr_model, pars = "pred_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ), 
  t, 
  mean_abundances,
  stddev_abundances
)
snc_usr_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_usr_model, pars = "pred_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ), 
  t, 
  mean_abundances,
  stddev_abundances
)
snc_sis_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sis_model, pars = "pred_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ), 
  t, 
  mean_abundances,
  stddev_abundances
)
snc_sis_ll_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sis_model, pars = "pred_ll_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ), 
  t, 
  mean_abundances,
  stddev_abundances
)
colnames(snc_sr_counts_pred) <- make.names(colnames(snc_usr_counts_pred)) # to remove % in the col names
colnames(snc_usr_counts_pred) <- make.names(colnames(snc_usr_counts_pred)) # to remove % in the col names
colnames(snc_sis_counts_pred) <- make.names(colnames(snc_sis_counts_pred)) # to remove % in the col names
colnames(snc_sis_ll_counts_pred) <- make.names(colnames(snc_sis_ll_counts_pred)) # to remove % in the col names


p_usr <- ggplot(snc_usr_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(data = snc_usr_counts_pred, aes(ymin = X5., ymax = X95.), fill = "blue", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances)) +
  geom_errorbar(
    data = snc_sr_counts_pred, 
    aes(
      ymin=mean_abundances - stddev_abundances,
      ymax=mean_abundances + stddev_abundances
    ),
    width=.2,
    position=position_dodge(0.05),
  ) +
  labs(title = "Unsaturated Removal Model Predictions", x = "Week", y = "SnC (normalized)", color = "name")

p_sr <- ggplot(snc_sr_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances)) +
  geom_errorbar(
    aes(
      ymin=mean_abundances - stddev_abundances,
      ymax=mean_abundances + stddev_abundances
    ),
    width=.2,
    position=position_dodge(0.05)
  ) +
  scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(8, 80)) +
  labs(title = "Saturated Removal Model Predictions", x = "Week", y = "SnC (normalized)", color = "name")
p_sis <- ggplot(snc_sis_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_line(data = snc_sis_ll_counts_pred, aes(x = t, y = X50.), col = "green") +
  geom_point(mapping = aes(y = mean_abundances)) +
  geom_errorbar(
    aes(
      ymin=mean_abundances - stddev_abundances,
      ymax=mean_abundances + stddev_abundances
    ),
    width=.2,
    position=position_dodge(0.05)
  ) +
  scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(8, 80)) +
  labs(title = "SIS Model Predictions", x = "Week", y = "SnC (normalized)", color = "name")
ggarrange(p_usr, p_sr, p_sis)
```


# Model comparison
## Bayes Factor and LOO-CV calculations
```{r}
if (params$refit_models) {
fit_snc_sr_model.bridge <- bridge_sampler(fit_snc_sr_model)
fit_snc_sis_model.bridge <- bridge_sampler(fit_snc_sis_model, silent = T)
fit_snc_usr_model.bridge <- bridge_sampler(fit_snc_usr_model, silent = T)
bayes_factor(fit_snc_sis_model.bridge, fit_snc_sr_model.bridge)
bayes_factor(fit_snc_sis_model.bridge, fit_snc_usr_model.bridge)
bayes_factor(fit_snc_sr_model.bridge, fit_snc_usr_model.bridge)
post_prob(fit_snc_sis_model.bridge, fit_snc_sr_model.bridge, fit_snc_usr_model.bridge)
}
```

```{r}
ll_sr <- extract_log_lik(fit_snc_sr_model, parameter_name= "log_likelihood")
rel_n_eff_sr <- relative_eff(exp(ll_sr), chain_id = rep(1:1, each = 4000))

loo(ll_sr, r_eff = rel_n_eff_sr)
rel_n_eff_sis <- relative_eff(exp(ll_sis), chain_id = rep(1:1, each = 4000))

ll_sis <- extract_log_lik(fit_snc_sis_model, parameter_name= "log_likelihood")
loo(ll_sis)
```

## Extrapolating in time
```{r}
n_weeks_is <- 8
n_weeks_oos <- length(weeks) - n_weeks_is

snc_abundance_is <- matrix(nrow = n_mice, ncol=n_weeks_is)
snc_abundance_oos <- matrix(nrow = n_mice, ncol=n_weeks_oos)

for (m in 1:n_mice) {
  for (t in 1:n_weeks_is) {
    mouse <- mice[m]
    week <- weeks[t]
    mouse_week_index <- which(snc_counts$mouse == mouse & snc_counts$week == week)[1]
    snc_abundance_is[m, t] <- snc_counts$norm_abundance[mouse_week_index]
  }
}
for (m in 1:n_mice) {
  for (t in (n_weeks_is+1):n_weeks) {
    mouse <- mice[m]
    week <- weeks[t]
    mouse_week_index <- which(snc_counts$mouse == mouse & snc_counts$week == week)[1]
    snc_abundance_oos[m, t - n_weeks_is] <- snc_counts$norm_abundance[mouse_week_index]
  }
}

# times
t_is <- sort(weeks)[1:n_weeks_is] # 8 hard-coded to match data
t_oos <- sort(weeks)[(n_weeks_is+1): n_weeks]

t0 = 0
t0_oos = n_weeks_is


# initial conditions
y0 <- c(1)
dim(y0) <- (1)
y0_oos <- mean_abundances[n_weeks_is];
dim(y0_oos) <- 1

# data for Stan
data_snc_split <- list(
  n_weeks = n_weeks_is, 
  n_mice = n_mice, 
  y0 = y0, 
  t0 = t0, 
  ts = t_is, 
  cells = snc_abundance_is,
  n_weeks_oos = n_weeks_oos,
  y0_oos = y0_oos,
  t0_oos = t0_oos
)
```

```{r}
sr_model <- stan_model("snc_sr_model.stan")
if (params$refit_models) {
  fit_snc_sr_split_model <- sampling(
    sr_model,
    data = data_snc_split,
    iter = niter,
    chains = 4,
    control = list(adapt_delta = .99)
  )
  saveRDS(fit_snc_sr_split_model, "saved_fit/fit_snc_sr_split_model.RDS")
} else {
  fit_snc_sr_split_model <- readRDS("saved_fit/fit_snc_sr_split_model.RDS")
}
```

```{r}
sis_model <- stan_model("snc_sis_model.stan")
if (params$refit_models) {
  fit_snc_sis_split_model <- sampling(
    sis_model,
    data = data_snc_split,
    iter = niter,
    chains = 4,
    control = list(adapt_delta = .99)
  )
  saveRDS(fit_snc_sis_split_model, "saved_fit/fit_snc_sis_split_model.RDS")
} else {
  fit_snc_sis_split_model <- readRDS("saved_fit/fit_snc_sis_split_model.RDS")
}
```
```{r}
sr_pars=c('eta', 'beta', 'sigma', 'lp__', 'log_likelihood')
print(fit_snc_sr_model, pars = sr_pars, digits_summary = 4)
print(fit_snc_sr_split_model, pars = sr_pars, digits_summary = 4)
```
```{r}
sis_pars=c('eta', 'beta', 'sigma', 'll', 'alpha', 'lp__', 'log_likelihood')
print(fit_snc_sis_model, pars = sis_pars)
print(fit_snc_sis_split_model, pars = sis_pars)
```

```{r}
snc_sr_oos_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sr_split_model, pars = "pred_cells_means_oos", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ),
  t = t_oos,
  mean_abundances = mean_abundances[(n_weeks_is+1): n_weeks],
  stddev_abundances = mean_abundances[(n_weeks_is+1): n_weeks]
)
snc_sis_oos_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sis_split_model, pars = "pred_cells_means_oos", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ),
  t = t_oos,
  mean_abundances = mean_abundances[(n_weeks_is+1): n_weeks],
  stddev_abundances = mean_abundances[(n_weeks_is+1): n_weeks]
)
colnames(snc_sr_oos_counts_pred) <- make.names(colnames(snc_sr_oos_counts_pred)) # to remove % in the col names
colnames(snc_sis_oos_counts_pred) <- make.names(colnames(snc_sis_oos_counts_pred)) # to remove % in the col names

ggplot(snc_sr_oos_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances)) +
  geom_errorbar(
    aes(
      ymin=mean_abundances - stddev_abundances,
      ymax=mean_abundances + stddev_abundances
    ),
    width=.2,
    position=position_dodge(0.05)
  ) +
  scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(t_oos[1], t_oos[n_weeks_oos])) +
  labs(title = "Saturated Removal Model Predictions", x = "Week", y = "SnC (normalized)", color = "name")
ggplot(snc_sis_oos_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances)) +
  geom_errorbar(
    aes(
      ymin=mean_abundances - stddev_abundances,
      ymax=mean_abundances + stddev_abundances
    ),
    width=.2,
    position=position_dodge(0.05)
  ) +
  scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(t_oos[1], t_oos[n_weeks_oos])) +
  labs(title = "SIS Model Predictions", x = "Week", y = "SnC (normalized)", color = "name")
```


# Perturbed Data
## Loading and visualizing the data
```{r}
snc_counts_fname <- str_c(
  params$data_dir, "raw", "senescent_cells_and_the_dynamics_of_aging__fig_3.csv",
  sep="/"
)
raw_perturbed_snc_counts <- read.csv(snc_counts_fname, header=TRUE)
long_pert_snc_counts <- raw_perturbed_snc_counts[
  (raw_perturbed_snc_counts$group == "BLEO_YNG") |
  (raw_perturbed_snc_counts$group == "BLEO_OLD"),
]
long_pert_snc_counts$day_after_treatment <- map_int(
  long_pert_snc_counts$day_after_treatment, 
  function (day) { return(as.integer(substr(day, 1, str_length(week)))) }
)
long_pert_snc_counts$week_after_treatment <- map_dbl(
  long_pert_snc_counts$day_after_treatment, 
  function (day) { return(as.numeric(as.integer(substr(day, 1, str_length(day))) / 7)) }
)
init_pert_snc_counts <- raw_perturbed_snc_counts[
  (raw_perturbed_snc_counts$group == "PBS_YNG") |
  (raw_perturbed_snc_counts$group == "PBS_OLD"),
]
t0_young <- 16
t0_old <- 94

init_pert_snc_counts
long_pert_snc_counts
```

```{r}
long_pert_snc_counts_young <- long_pert_snc_counts[long_pert_snc_counts$group == "BLEO_YNG",]
long_pert_snc_counts_old <- long_pert_snc_counts[long_pert_snc_counts$group == "BLEO_OLD",]

mean_abundances_perturb_young <- aggregate(
  long_pert_snc_counts_young$SnC, 
  list(long_pert_snc_counts_young$week_after_treatment), 
  median, na.rm = T
)$x
mean_abundances_perturb_old <- aggregate(
  long_pert_snc_counts_old$SnC, 
  list(long_pert_snc_counts_old$week_after_treatment), 
  median, na.rm = T
)$x
stddev_abundances_perturb_young <- aggregate(
  long_pert_snc_counts_young$SnC, 
  list(long_pert_snc_counts_young$week_after_treatment), 
  sd, na.rm = T
)$x
stddev_abundances_perturb_old <- aggregate(
  long_pert_snc_counts_old$SnC, 
  list(long_pert_snc_counts_old$week_after_treatment), 
  sd, na.rm = T
)$x

mean_stddev_abundances_perturb_young <- data.frame(
  mean = mean_abundances_perturb_young,
  stddev = stddev_abundances_perturb_young,
  week = unique(long_pert_snc_counts_young$week_after_treatment)
)
mean_stddev_abundances_perturb_young <- data.frame(
  mean = mean_abundances_perturb_young,
  stddev = stddev_abundances_perturb_young,
  week = unique(long_pert_snc_counts_young$week_after_treatment)
)
p <- ggplot(mean_stddev_abundances_perturb_young, aes(x=week, y=mean)) + geom_point() +
    geom_errorbar(aes(ymin=mean-stddev/2, ymax=mean+stddev/2), width=.2,
                 position=position_dodge(0.05))
p
```


```{r}
ggplot(
  data = long_pert_snc_counts, 
  aes(x = day_after_treatment, y = SnC, color = group)
) +
  xlab("Day After Treatment") + ylab("SnC Abundance (Normalized)") +
  geom_point() +
  scale_y_continuous(trans = log_trans(), breaks = c(0.1, 1, 5, 20, 50))
```

## Half-life calculation
```{r}
beta_sr_summary <- summary(
  fit_snc_sr_model, 
  pars = c("beta")
)$summary

# Young
beta_sr_mean <- beta_sr_summary[, c("mean")]
half_life_sr_mean_young <- 1 / (
  mean(beta_sr_mean[1] / (1 / beta_sr_mean[2] + init_pert_snc_counts$SnC[init_pert_snc_counts$group == "PBS_YNG"])) * (log(2)) ** (-1)
)
half_life_sr_mean_young

# Old
beta_sr_mean <- beta_sr_summary[, c("mean")]
half_life_sr_mean_old <- 1 / (
  mean(beta_sr_mean[1] / (1 / beta_sr_mean[2] + init_pert_snc_counts$SnC[init_pert_snc_counts$group == "PBS_OLD"])) * (log(2)) ** (-1)
)
half_life_sr_mean_old
```
## Predicted LL SnCs at old & young start times
```{r}
snc_sis_ll_counts_pred_young <- 
  as.data.frame(
    summary(
      fit_snc_sis_model, pars = "pred_ll_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  )
snc_sis_ll_counts_pred_old <- 
  as.data.frame(
    summary(
      fit_snc_sis_model, pars = "pred_ll_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  )
t
snc_sis_ll_counts_pred_young
snc_sis_ll_counts_pred_old
```
## Predicting pulse data results
```{r}
t0_young <- 16
t0_old <- 94

ll_young <- c(mean = 0.4228642, sd = .2310055)
ll_old <- c(mean = 1.6076111, sd = 0.9887867)

# initial conditions for both cases
y0_young <- mean(init_pert_snc_counts$SnC[init_pert_snc_counts$group == "PBS_YNG"]) + 50

dim(y0_young) <- 1
y0_old <- mean(init_pert_snc_counts$SnC[init_pert_snc_counts$group == "PBS_OLD"]) + 50
dim(y0_old) <- 1

beta_sr_param <- summary(
  fit_snc_sr_model, 
  pars = c("beta[1]", "beta[2]")
)$summary[, c("mean")]


# data for Stan
snc_abundance_young <- data.frame(long_pert_snc_counts[
  long_pert_snc_counts$group == "BLEO_YNG",
])
weeks_young <- unique(snc_abundance_young$week_after_treatment)
t_young <- weeks_young + t0_young
n_weeks_young <- length(weeks_young)
cells_young <- snc_abundance_young$SnC
tt_young <- match(snc_abundance_young$week_after_treatment, weeks_young)
snc_abundance_old <- data.frame(long_pert_snc_counts[
  long_pert_snc_counts$group == "BLEO_OLD",
])
weeks_old <- unique(snc_abundance_old$week_after_treatment)
n_weeks_old <- length(weeks_old)
n_mice_young <- 0
n_mice_old <- 0
for (w in weeks_young) {
  counts <- snc_abundance_young$SnC[snc_abundance_young$week_after_treatment == w]
  n_mice_young <- max(n_mice_young, length(counts))
}
for (w in weeks_old) {
  counts <- snc_abundance_old$SnC[snc_abundance_old$week_after_treatment == w]
  n_mice_old <- max(n_mice_old, length(counts))
}

t_old <- weeks_old + t0_old
cells_old <- snc_abundance_old$SnC
tt_old <- match(snc_abundance_old$week_after_treatment, weeks_old)
data_snc_sr_young <- list(
  n_weeks = n_weeks_young,
  n_mice = n_mice_young,
  cells = cells_young, 
  n_obs = length(cells_young), 
  tt = tt_young, 
  y0 = y0_young, 
  t0 = t0_young, 
  ts = t_young, 
  beta = beta_sr_param
)
data_snc_sr_old <- list(
  n_weeks = n_weeks_old,
  n_mice = n_mice_old,
  cells = cells_old,
  n_obs = length(cells_old),
  tt = tt_old,
  y0 = y0_old,
  t0 = t0_old,
  ts = t_old,
  beta = beta_sr_param
)

# SIS stan pre-processing
beta_sis_param <- summary(
  fit_snc_sis_model, 
  pars = c("beta[1]", "beta[2]")
)$summary[, c("mean")]
alpha_sis_param <- summary(
  fit_snc_sis_model,
  pars = c("alpha")
)$summary[, c("mean")]
# data for Stan
data_snc_sis_young <- list(
  n_weeks = n_weeks_young, 
  n_mice = n_mice_young,
  cells = cells_young, 
  n_obs = length(cells_young), 
  tt = tt_young, 
  y0 = y0_young, 
  t0 = t0_young, 
  ts = t_young, 
  beta = beta_sis_param,
  ll_param = ll_young
)
data_snc_sis_old <- list(
  n_weeks = n_weeks_old,
  cells = cells_old,
  n_mice = n_mice_old,
  n_obs = length(cells_old),
  tt = tt_old,
  y0 = y0_old,
  t0 = t0_old,
  ts = t_old,
  beta = beta_sis_param,
  ll_param = ll_old
)

# number of MCMC steps
niter <- 4000
```



```{r}
snc_sr_sim_model <- stan_model("snc_sr_model_pulse_sim.stan")
fit_snc_sr_sim_young <- sampling(
  snc_sr_sim_model,
  data = data_snc_sr_young,
  iter = niter,
  chains = 4,
  control = list(adapt_delta = .99)
)
fit_snc_sr_sim_old <- sampling(
  snc_sr_sim_model,
  data = data_snc_sr_old,
  iter = niter,
  chains = 4,
  control = list(adapt_delta = .99)
)
```

```{r}
snc_sis_sim_model <- stan_model("snc_sis_model_pulse_sim.stan")
fit_snc_sis_sim_young <- sampling(
  snc_sis_sim_model,
  data = data_snc_sis_young,
  iter = niter,
  chains = 4,
  control = list(adapt_delta = .99)
)
fit_snc_sis_sim_old <- sampling(
  snc_sis_sim_model,
  data = data_snc_sis_old,
  iter = niter,
  chains = 4,
  control = list(adapt_delta = .99)
)
```

```{r}
sr_pars=c('eta', 'sigma', 'beta', 'lp__')
print(fit_snc_sr_sim_young, pars = sr_pars, digits_summary = 4)
print(fit_snc_sr_sim_old, pars = sr_pars, digits_summary = 4)
```


```{r}
sis_pars <- c('eta', 'll', 'beta', 'sigma', 'lp__')
print(fit_snc_sis_sim_young, pars = sis_pars, digits_summary = 4)
print(fit_snc_sis_sim_old, pars = sis_pars, digits_summary = 4)
```

```{r}
pairs(fit_snc_sr_sim_old, pars=sr_pars)
```

```{r}
snc_sr_sim_counts_pred_young <- cbind(
  as.data.frame(
    summary(
      fit_snc_sr_sim_young, pars = "pred_cells", 
      probs = c(0.25, 0.5, 0.75)
    )$summary
  ), 
  t_young,
  mean_abundances_perturb_young,
  stddev_abundances_perturb_young
)
snc_sr_sim_counts_pred_old <- cbind(
  as.data.frame(
    summary(
      fit_snc_sr_sim_old, pars = "pred_cells", 
      probs = c(0.25, 0.5, 0.75)
    )$summary
  ), 
  t_old,
  mean_abundances_perturb_old,
  stddev_abundances_perturb_old
)
snc_sis_sim_counts_pred_young <- cbind(
  as.data.frame(
    summary(
      fit_snc_sis_sim_young, pars = "pred_cells", 
      probs = c(0.25, 0.5, 0.75)
    )$summary
  ), 
  t_young,
  mean_abundances_perturb_young,
  stddev_abundances_perturb_young
)
snc_sis_sim_counts_pred_old <- cbind(
  as.data.frame(
    summary(
      fit_snc_sis_sim_old, pars = "pred_cells", 
      probs = c(0.25, 0.5, 0.75)
    )$summary
  ), 
  t_old,
  mean_abundances_perturb_old,
  stddev_abundances_perturb_old
)
colnames(snc_sr_sim_counts_pred_young) <- make.names(colnames(snc_sr_sim_counts_pred_young)) # to remove % in the col names
colnames(snc_sr_sim_counts_pred_old) <- make.names(colnames(snc_sr_sim_counts_pred_old)) # to remove % in the col names
colnames(snc_sis_sim_counts_pred_young) <- make.names(colnames(snc_sis_sim_counts_pred_young)) # to remove % in the col names
colnames(snc_sis_sim_counts_pred_old) <- make.names(colnames(snc_sis_sim_counts_pred_old)) # to remove % in the col names

p_sr_sim_young <- ggplot(snc_sr_sim_counts_pred_young, mapping = aes(x = t_young)) +
  geom_ribbon(aes( ymin = X25., ymax = X75.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t_young, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances_perturb_young)) +
  scale_x_continuous(name = "Week", breaks=t_young, labels=scales::number_format(accuracy = 0.1), limits = c(min(t_young), max(t_young))) +
  labs(title = "Saturated Removal Model Predictions (Young)", x = "Week", y = "SnC (normalized)", color = "name")
p_sr_sim_old <- ggplot(snc_sr_sim_counts_pred_old, mapping = aes(x = t_old)) +
  geom_ribbon(aes( ymin = X25., ymax = X75.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t_old, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances_perturb_old)) +
  scale_x_continuous(name = "Week", breaks=t_old, labels=scales::number_format(accuracy = 0.1), limits = c(min(t_old), max(t_old))) +
  labs(title = "Saturated Removal Model Predictions (Old)", x = "Week", y = "SnC (normalized)", color = "name")
p_sis_sim_young <- ggplot(snc_sis_sim_counts_pred_young, mapping = aes(x = t_young)) +
  geom_ribbon(aes( ymin = X25., ymax = X75.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t_young, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances_perturb_young)) +
  scale_x_continuous(name = "Week", breaks=t_young, labels=scales::number_format(accuracy = 0.1), limits = c(min(t_young), max(t_young))) +
  labs(title = "Saturated Removal Model Predictions (Young)", x = "Week", y = "SnC (normalized)", color = "name")
p_sis_sim_old <- ggplot(snc_sis_sim_counts_pred_old, mapping = aes(x = t_old)) +
  geom_ribbon(aes( ymin = X25., ymax = X75.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t_old, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances_perturb_old)) +
  scale_x_continuous(name = "Week", breaks=t_old, labels=scales::number_format(accuracy = 0.1), limits = c(min(t_old), max(t_old))) +
  labs(title = "Saturated Removal Model Predictions (Old)", x = "Week", y = "SnC (normalized)", color = "name")

ggarrange(p_sr_sim_young, p_sr_sim_old, p_sis_sim_young, p_sis_sim_old)
```
## Model Comparison
```{r}
fit_snc_sr_sim_young.bridge <- bridge_sampler(fit_snc_sr_sim_young)
fit_snc_sr_sim_old.bridge <- bridge_sampler(fit_snc_sr_sim_old)
fit_snc_sis_sim_young.bridge <- bridge_sampler(fit_snc_sis_sim_young)
fit_snc_sis_sim_old.bridge <- bridge_sampler(fit_snc_sis_sim_old)

bayes_factor(fit_snc_sis_sim_young.bridge, fit_snc_sr_sim_young.bridge)
bayes_factor(fit_snc_sis_sim_old.bridge, fit_snc_sr_sim_old.bridge)
```
# Hierarchical Model
```{r}
# fit_snc_sr_hier_model <- stan(
#   file = "snc_sr_model_hierarchical.stan",
#   data = data_snc,
#   iter = niter * 2,
#   chains = 4,
#   control = list(max_treedepth = 15, adapt_delta = .95)
# )
# saveRDS(
#   fit_snc_sr_hier_model, 
#   "saved_fit/fit_snc_sr_model_hierarchical.RDS"
# )
```

```{r}
# sr_hierarchical_pars=c('mu', 'tau', 'sigma', 'lp__', 'log_likelihood')
# print(fit_snc_sr_hier_model, pars = sr_hierarchical_pars, digits_summary = 4)
```


```{r}
# pairs(fit_snc_sr_hier_model, pars=c("sigma", "tau", "mu"))
```

```{r}
# snc_sr_hier_counts_pred <- cbind(
#   as.data.frame(
#     summary(
#       fit_snc_sr_hier_model, pars = "pred_cells", 
#       probs = c(0.05, 0.5, 0.95)
#     )$summary
#   ), 
#   t,
#   mouse = sort(rep(1:n_mice, times = n_weeks))
# )
# colnames(snc_sr_hier_counts_pred) <- make.names(colnames(snc_sr_hier_counts_pred)) # to remove % in the col names
# plots <- c()
# make_pp_plot <- function(i) {
#   mice_i_pred_counts <- subset(snc_sr_hier_counts_pred, mouse == i)
#   mice_i_real_counts <- snc_abundance[i,]
#   p <- ggplot(mice_i_pred_counts, mapping = aes(x = t)) +
#     geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
#     geom_line(mapping = aes(x = t, y = X50.)) +
#     geom_point(mapping = aes(y = mice_i_real_counts)) +
#     scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(8, 80)) +
#     labs( x = "Week", y = "SnC (normalized)", color = "name")
#     # facet_wrap()
#   return(p)
# }
# ggarrange(plotlist = lapply(1:n_mice, make_pp_plot), ncol = 3, nrow = 2,
#           common.legend = T)
```


