---
title: "Towards validation of the senescence-induced senescence hypothesis"
author: "Stephen Malina"
output:
  bookdown::pdf_document2: 
    toc: false
  bookdown::html_document2: default
  extra_dependencies: ["float"]
bibliography: bib.bib
link-citations: yes
---

```{r include=FALSE}
library(bookdown)
library(brms)
library(dplyr)
library(foreach)
library(ggplot2)
library(ggpubr)
library(latex2exp)
library(loo)
library(purrr)
library(reshape2)
library(stringr)
library(tidyverse)

library(rmarkdown)
library(knitr)

# Stan-specific stuff
library(bridgesampling)
library(rstan)
library(gridExtra)
rstan_options (auto_write = TRUE)
options (mc.cores = parallel::detectCores ())
set.seed(3) # for reproductibility
options(digits = 2)
```

# Introduction
Cellular senescence is increasingly believed to play an important role in the aging process.
Senescent cells (SnCs) are cells that can no longer divide but continue to limp along
rather than actually die. In young, healthy organisms, senescent
cells can play a positive role in arresting tumor growth and healing wounds
[@ricon2020senescence; @he2017senescence; @krtolica2001senescent]. 
However, increasing evidence points to senescent cells contributing to the aging
process via slowing stem cell growth and tissue regeneration and increasing
inflammation in older organisms [@campisi2007cellular].

A characteristic pattern that accompanies
the shift in roles is an increase in relative abundance of and slowing down
of turnover of senescent cells. In order
to understand and intervene on cellular senescence to prolong healthy lifespan
and ward off disease, we must understand the factors that drive the shift
SnC growth regimes. In particular, one relative under-explored hypothesis 
from @nelson2012senescent suggests
that the SnC turn over slow down may partially result from a "bystander effect"
in which senescentcells induce nearby healthy cells into senescence 
(senescence-induced senescence, SIS).

We leveraged longitudinal and
pulse data [@karin2019senescent; @burd2013monitoring] to test
how well a model that includes SIS from long-lived senescent cells
captured observed senescent cell dynamics relative to two already published
models [@karin2019senescent]. We also computed a point estimate and bounds on the fraction
of senescent cells which our models predicted to be long-lived based on
longitudinal data. 

# Methods
We built a Bayesian ODE-based model of SnC growth dynamics based on a pair
of mouse SnC time-series datasets using stan [@carpenter2017stan]. We compare
our model to Bayesian versions of two previously-published models using
Bayes Factors [@bridgesampling], Leave-One-Out Cross-Validation (LOO-CV) [@loo-cv],
and an extrapolation test. All of our code and data are available on Github at
https://github.com/an1lam/sis.

## Datasets
### Longitudinal Dataset
```{r echo = F}
snc_counts <- readRDS("saved_fit/snc_counts.RDS")
n_mice <- max(snc_counts$mouse)
max_week <- max(snc_counts$week)
```

Our primary dataset (originally collected by @burd2013monitoring) measures SnC
abundance in `r n_mice` mice over their entire
lifespan (`r max_week` weeks, i.e. ~`r max_week / 52` years). This dataset
consists of total body luminescence (TBL)
measurements from each mouse taken at 2 week intervals throughout each mouse's
life. These luminescence measurements
act as a proxy for \( \text{p16}^{\text{INK4a}} \), a senescence biomarker, 
prevalence. \( \text{p16}^{\text{INK4a}} \) itself then serves
as a proxy for senescent cell prevalence throughout the body. In line
with @karin2019senescent, we preprocess unnormalized TBL values into normalized
SnC abundance measurements (AUs) by dividing all luminescence values 
by mean luminescence during the first three weeks of the experiment (weeks 8, 16, 24).

### Pulse Dataset
The SnC pulse dataset (from @karin2019senescent) measures SnC abundance in old 
(22 months) and young (3 months) mices' lungs
after treatment with bleomycin, a compound known to induce cellular senescence
a few days after application. Similar to as with the longitudinal dataset,
all SnC abundance measurements are normalized to the 

## Models
## Bayesian Saturated Removal (SR) and Unsaturated Removal (USR) base models
As a first step towards building and assessing a model that includes
senescence-induced senescence, we re-fit two existing models [@karin2019senescent]
of senescent cell growth on the longitudinal senescent cell dataset.

In their paper,  @karin2019senescent apply a maximum likelihood framework and parameter
grid search to fit 16 ODE-based models of senescent cell growth. Each of
their 16 models was derived by selecting a subset of parameters from the
following 'full' model of senescent cell growth with time/age,

\begin{equation}
\frac{\rm{d}C}{\rm{d}t} = \underbrace{(\eta_0 + \eta_1 * t)(1 + \eta_2 * C)}_{\text{production}}
- \underbrace{\frac{\beta_0 - \beta_1 * t}{1 + \beta_2 * C} * C}_{\text{removal}} + 
\underbrace{\sqrt{2 \epsilon} * \xi_t}_{\text{noise}}.
  (\#eq:1)
\end{equation}

In equation \@ref(eq:1), \( C \) denotes the normalized senescent cell abundance, measured
in arbitrary units (AUs) (see Appendix for details). 
The first of three expression's three variables,
\( \eta_0,  \eta_1, \eta_2 \),
denote initial growth rate, growth rate increase with age, and
senescence-induced senescence rate respectively. The second of the three expression's
three variables, \( \beta_0, \beta_1, \beta_2 \), denote initial removal rate,
decrease in removal rate with age, and inverse of the half-way saturation
point for senescent cell removal respectively. Last, the remaining term
\( \xi_t \) models per time-step independent noise, which we replace
with a single independent noise term.

In @karin2019senescent, the best-performing model variant, the saturated removal (SR) model,
got rid of the senescence-induced senescence term but otherwise left the
model as-is. The best-performing variant without the saturated removal term (\( \beta_2 \)), i.e. the unsaturated removal (USR) model,
also excluded the senescence-induced senescence term in its growth rate 
expression.

```{r fig-1, echo = F, fig.cap = "Unnormalized senescent cell (SnC) abundance trajectories for each mouse (top left). Retrodictions and 5%/95% credible intervals for SnC mean normalized abundance predicted by SR, USR, and SIS models (top right, bottom left, bottom right) respectively. Lines denote posterior median, dots denote observed mean abundance with error bars representing sample standard error. Green line in the SIS graph denotes estimated long-lived SnC abundance.", fig.height = 3}

# Load relevant objects
data_snc <- readRDS("saved_fit/data_snc.RDS")
snc_counts <- readRDS("saved_fit/snc_counts.RDS")
fit_snc_sr_model <- readRDS("saved_fit/fit_snc_sr_model.RDS")
fit_snc_usr_model <- readRDS("saved_fit/fit_snc_usr_model.RDS")
fit_snc_sis_model <- readRDS("saved_fit/snc_sis_model.RDS")

t <- data_snc$ts
mean_abundances <- aggregate(
  snc_counts$norm_abundance, list(snc_counts$week), mean, na.rm = T
)$x
stddev_abundances <- aggregate(
  snc_counts$norm_abundance, list(snc_counts$week), sd, na.rm = T
)$x

snc_sr_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sr_model, pars = "pred_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ), 
  t, 
  mean_abundances,
  stddev_abundances
)
snc_usr_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_usr_model, pars = "pred_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ), 
  t, 
  mean_abundances,
  stddev_abundances
)
snc_sis_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sis_model, pars = "pred_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ), 
  t, 
  mean_abundances,
  stddev_abundances
)
snc_sis_ll_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sis_model, pars = "pred_ll_cells_means", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ), 
  t, 
  mean_abundances,
  stddev_abundances
)
colnames(snc_sr_counts_pred) <- make.names(colnames(snc_usr_counts_pred)) # to remove % in the col names
colnames(snc_usr_counts_pred) <- make.names(colnames(snc_usr_counts_pred)) # to remove % in the col names
colnames(snc_sis_counts_pred) <- make.names(colnames(snc_sis_counts_pred)) # to remove % in the col names
colnames(snc_sis_ll_counts_pred) <- make.names(colnames(snc_sis_ll_counts_pred)) # to remove % in the col names


p_usr <- ggplot(snc_usr_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(data = snc_usr_counts_pred, aes(ymin = X5., ymax = X95.), fill = "blue", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances)) +
  geom_errorbar(
    data = snc_sr_counts_pred, 
    aes(
      ymin=mean_abundances - stddev_abundances,
      ymax=mean_abundances + stddev_abundances
    ),
    width=.2,
    position=position_dodge(0.05),
  ) +
  labs(title = "USR Model Predictions", x = "Week", y = "SnC (normalized)", color = "name")

p_sr <- ggplot(snc_sr_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances)) +
  geom_errorbar(
    aes(
      ymin=mean_abundances - stddev_abundances,
      ymax=mean_abundances + stddev_abundances
    ),
    width=.2,
    position=position_dodge(0.05)
  ) +
  scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(8, 80)) +
  labs(title = "SR Model Predictions", x = "Week", y = "SnC (normalized)", color = "name")
p_sis <- ggplot(snc_sis_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_line(data = snc_sis_ll_counts_pred, aes(x = t, y = X50.), col = "green") +
  geom_point(mapping = aes(y = mean_abundances)) +
  geom_errorbar(
    aes(
      ymin=mean_abundances - stddev_abundances,
      ymax=mean_abundances + stddev_abundances
    ),
    width=.2,
    position=position_dodge(0.05)
  ) +
  scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(8, 80)) +
  labs(title = "SIS Model Predictions", x = "Week", y = "SnC (normalized)", color = "name")

p_raw <- ggplot(
  data = snc_counts, 
  aes(x = week, y = TBL_x_10_5, color = mouse, group = mouse),
  show.legend = F
) +
  labs(x = "Week", y = TeX("$ TBL \\times 10^5 $"), title = "SnC Abundance by Mouse") +
  scale_color_gradientn(colours = rainbow(5)) + geom_line(show.legend = F) + geom_point(show.legend = F)

grid.arrange(p_raw, p_usr, p_sr, p_sis, layout_matrix = matrix(c(1, 2 , 3, 4), nrow = 2))
```

As a baseline
for comparing our modified senescence-induced senescence model, we reproduced
@karin2019senescent's SR and USR models as Bayesian models in stan [@carpenter2017stan] 
(see Appendix \@ref(model-diffs) for details on priors, noise model, and stan configuration).
At a high level, our results matched theirs. We compared the posterior probability
of the two models and found that the SR model was 50.3 times more likely than
the USR model (@bridgesampling).

Although the SR model's full posterior fits
the data meaningfully better than the USR model's, as Figure \@ref(fig:fig-1) 
shows, both models' credible intervals and posterior means mostly capture the 
observed data's evolution through time. This suggests that the maximum likelihood
approach may have underestimated the quality of the USR model even if it
correctly captures the superiority of the SR model.

### Long-lived SnC senescence-induced senescence (SIS) model 
As mentioned, the original full model included a term for SIS. 
However, the model treated SIS as growing as a function of overall
SnC count and total SnC growth rate. Intuitively, given a long enough time
this would inevitably lead to exponential growth of the number of senescent cells,
which we don't observe. As suggested by @van2014role, it's instead possible
that a small fraction of senescent cells both turn over more slowly and
induce senescence in their neighbors, but the majority do not.

To test this hypothesis, we can modify\footnote{As discussed in the appendix, we also remove the independent
noise and instead model it as a noise-variance term in our likelihood.}the SR model by replacing
the current time-dependent growth rate term with a time-dependent
long-lived cell SIS term:

\begin{align}
\frac{\rm{d}L}{\rm{d}t} &= \alpha \\
\frac{\rm{d}C}{\rm{d}t} &= (\eta_0 + \eta_2 * L) -
  \frac{\beta_0 - \beta_1 * t}{1 + \beta_2 * C}.
(\#eq:2)
\end{align}

In equation \@ref(eq:2), \( \frac{\rm{d}L}{\rm{d}t} \) denotes the time rate of change of
(latent) long-lived senescent cell count. Solving for \( L \) this becomes
\begin{equation}
\frac{\rm{d}C}{\rm{d}t} = (\eta_0 + \eta_2 * l + \eta_2 * \alpha t) - 
  \frac{\beta_0 - \beta_1 * t}{1 + \beta_2 * C} * C.
(\#eq:3)
\end{equation}
where \( l \) denotes the long-lived SnC abundance and \( \alpha t \) the
(assumed to be linear) increase as a function of time.

### Modifications for pulse dataset
In order to compare models in a second regime,
we fit variants of SR and SIS models on the pulse dataset. Given this dataset's
small number of mice and time points (5 weeks in the young mice cohort, 2
weeks in the old mice cohort), we used our posterior estimates of parameters
from the longitudinal model as priors for the corresponding parameters
in the pulse models. We also removed the time-dependent term from the
SIS model after finding it didn't contribute to the quality of the model's
fit.

# Results
## Long-lived SnC SIS model fits the longitudinal dataset better than the base models
Our new long-lived SnC SIS model accurately describes the SnC growth dynamics
of the longitudinal dataset. In fact, we find that it fits the data, as measured by
both Bayes Factor and LOO-CV, better (under a uniform prior between models for the Bayes Factor) than
the SR model (Bayes Factor of 25 and slightly higher ELPD-LOO). 
This suggests that long-lived senescent cell SIS may partially
explain senescent cell dynamics when combined with saturated removal.

```{r fig-2, echo = F, fig.cap = "SR and SIS model predictive distributions for held-out weeks 72 and 80. As in Figure 1, dots denote true abundance values, lines predictive medians, and yellow bars 5%/95% predictive intervals. Both models incorrectly predict a slowdown in SnC growth but capture both points in their predictive distributions.", fig.height = 2}

# Load relevant objects
data_snc <- readRDS("saved_fit/data_snc.RDS")
snc_counts <- readRDS("saved_fit/snc_counts.RDS")
fit_snc_sr_split_model <- readRDS("saved_fit/fit_snc_sr_split_model.RDS")
fit_snc_sis_split_model <- readRDS("saved_fit/fit_snc_sis_split_model.RDS")

t <- data_snc$ts
mean_abundances <- aggregate(
  snc_counts$norm_abundance, list(snc_counts$week), mean, na.rm = T
)$x
stddev_abundances <- aggregate(
  snc_counts$norm_abundance, list(snc_counts$week), sd, na.rm = T
)$x

# weeks & times
mice <- unique(snc_counts$mouse)
weeks <- unique(snc_counts$week)
n_weeks <- length(weeks)
n_weeks_is <- 8
n_weeks_oos <- n_weeks - n_weeks_is
t_is <- sort(weeks)[1:n_weeks_is] # 8 hard-coded to match data
t_oos <- sort(weeks)[(n_weeks_is+1): n_weeks]

snc_sr_oos_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sr_split_model, pars = "pred_cells_means_oos", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ),
  t = t_oos,
  mean_abundances = mean_abundances[(n_weeks_is+1): n_weeks],
  stddev_abundances = mean_abundances[(n_weeks_is+1): n_weeks]
)
snc_sis_oos_counts_pred <- cbind(
  as.data.frame(
    summary(
      fit_snc_sis_split_model, pars = "pred_cells_means_oos", 
      probs = c(0.05, 0.5, 0.95)
    )$summary
  ),
  t = t_oos,
  mean_abundances = mean_abundances[(n_weeks_is+1): n_weeks],
  stddev_abundances = mean_abundances[(n_weeks_is+1): n_weeks]
)

colnames(snc_sr_oos_counts_pred) <- make.names(colnames(snc_sr_oos_counts_pred)) # to remove % in the col names
colnames(snc_sis_oos_counts_pred) <- make.names(colnames(snc_sis_oos_counts_pred)) # to remove % in the col names

p_sr_split <- ggplot(snc_sr_oos_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances)) +
  scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(t_oos[1], t_oos[n_weeks_oos])) +
  labs(title = "SR Model Extrapolation", x = "Week", y = "SnC (normalized)", color = "name")
p_sis_split <- ggplot(snc_sis_oos_counts_pred, mapping = aes(x = t)) +
  geom_ribbon(aes( ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(x = t, y = X50.)) + 
  geom_point(mapping = aes(y = mean_abundances)) +
  scale_x_continuous(name = "Week", breaks=t, labels=t, limits = c(t_oos[1], t_oos[n_weeks_oos])) +
  labs(title = "SIS Model Extrapolation", x = "Week", y = "SnC (normalized)", color = "name")
ggarrange(p_sr_split, p_sis_split)
```

That said, when we compare this model's retrodictions of mean SnC count to the SR
model's, we can see essentially no difference between the two (again see Figure \@ref(fig:fig-1)).
Given this, to try and further validate our claim that the SIS model is better, we
tested the SR and SIS models' ability to extrapolate to held-out time steps. 
We split the SnC abundance observations into train and test set with the train set
consisting of the first 7 weeks of abundance observations and the test set the
final 3. Unfortunately, as \@ref(fig:fig-2) shows, neither model performed particularly well at extrapolation with both
wrongly predicting a decline in SnC growth. Taking this poor performance on an
absolute scale into account, we see that the SIS model slightly outperformed the
SR model insofar as its upper error bar comes closer to correctly predicting
the continued growth in SnC count. We hypothesize that this poor performance
results from the final 2 weeks' worth of data playing an important role
in helping the model learn the saturation points.
This belief is bolstered by the fact that both
the SR and SIS models trained on only the first 8 weeks of SnC abundance
data learn higher saturation points than the corresponding
full models but basically identical growth rates and long-lived SnC counts.

## Long-lived SnC SIS model doesn't fit the pulse data as well as the SR model
Unlike with the longitudinal dataset, on the pulse dataset, the SIS model's
increased complexity does not seem to improve the model's fit. As measured
by Bayes Factor, the SR model is approximately twice as probable as the SIS
model with respect to the pulse data. We admittedly didn't predict this
before running the model. In retrospect, it can be partially explained by a combination of
the pulse dataset having been generated in a way that likely would only
increase the number of acute SnCs, not long-lived ones, and the long-lived
cell SIS not contributing meaningful growth in the relatively short timescale
of the pulse dataset. In spite of this justification, this finding
should be taken as providing some, albeit weak evidence, against the long-lived
SIS hypothesis.

## SIS model estimates that long-lived SnCs comprise between 10% and 40% of all SnCs
```{r echo = F}
options(digits=3)
snc_counts <- readRDS("saved_fit/snc_counts.RDS")
fit_snc_sis_model <- readRDS("saved_fit/snc_sis_model.RDS")
pred_ll_cells_summary <- summary(
  fit_snc_sis_model,
  pars = c("pred_ll_cells_means", "ll"),
  probs = c(.05, .5, .95)
)$summary
colnames(pred_ll_cells_summary) <- make.names(colnames(pred_ll_cells_summary)) # to remove % in the col names

# pred_ll_cells_summary["pred_ll_cells_means[10]", "X50."] / snc_counts$norm_abundance[10]
mean_start_norm_abundance <- mean(
  snc_counts$norm_abundance[which(snc_counts$week == min(snc_counts$week))]
)
mean_final_norm_abundance <- mean(
  snc_counts$norm_abundance[which(snc_counts$week == max(snc_counts$week))]
)
ll_cells_start_frac_median <- pred_ll_cells_summary["pred_ll_cells_means[1]", "X50."] / mean_start_norm_abundance
ll_cells_final_frac_median <- pred_ll_cells_summary["pred_ll_cells_means[10]", "X50."] / mean_final_norm_abundance
ll_cells_start_frac_lower <- pred_ll_cells_summary["pred_ll_cells_means[1]", "X5."] / mean_start_norm_abundance
ll_cells_final_frac_lower <- pred_ll_cells_summary["pred_ll_cells_means[10]", "X5."] / mean_final_norm_abundance
ll_cells_start_frac_upper <- pred_ll_cells_summary["pred_ll_cells_means[1]", "X95."] / mean_start_norm_abundance
ll_cells_final_frac_upper <- pred_ll_cells_summary["pred_ll_cells_means[10]", "X95."] / mean_final_norm_abundance
```

Our model's median estimate for fraction of SnCs that are long-lived starts at
`r ll_cells_start_frac_median` and grows to a final value of `r ll_cells_final_frac_median`,
with 5%/95% lower and upper bounds of `r ll_cells_start_frac_lower`/`r ll_cells_start_frac_upper` and 
`r ll_cells_final_frac_lower`/`r ll_cells_final_frac_upper`. Keeping in mind
that our model infers its parameters from a small, noisy dataset, this
implies that long-lived SnCs could account for a meaningful fraction of
all SnCs, with this fraction growing dramatically throughout the mouse
lifespan.

# Discussion
## Strengths and Limitations
Our work has a few strengths. First, by using a fully Bayesian approach, 
we're better able to quantify our uncertainty about different parameter values.
This directly enabled us to, for example, recognize that our SIS model
put fairly wide bounds on the relative abundance of long-lived senescent
cells. By employing stan and its suite of diagnostic tools, we were also
better able to verify our model's fit and verify its parameter estimates
satisfied various sanity checks. 
Finally, the Bayesian approach enables more sophisticated model comparison,
which showed both the better fit of the SIS model on the longitudinal data
and the marginally worse fit of it on the pulse data.

On the other hand, our work is limited by our datasets and our modeling assumptions.
The longitudinal dataset we use has three major limitations:
\begin{enumerate}
\item It's relatively small, leading to wide error bars on our parameter estimates and
  also preventing us from accurately modeling inter-mouse heterogeneity.
\item It lacks spatial information, preventing us from testing whether SIS
  is mediated by cellular proximity.
\item Its time steps are coarse-grained relative to a mouse's lifespan, making
  extrapolation difficult.
\end{enumerate}
The pulse dataset is even smaller than the longitudinal dataset and
this made extracting useful additional information on parameter estimates
from it a challenge, especially in the case of the old mouse cohort.

On the modeling side, our SIS model is partially unidentifiable because
increases in \( \eta_2 \) can be partially compensated by decreases in \( \alpha \)
and vice versa. Fundamentally, this is necessary as long as the number of 
long-lived cells are latent but it complicates inference nonetheless.

## Implications
While our results are not strong enough to enable a definitive conclusion
regarding long-lived SnCs' role in shifting
SnC growth dynamics, they suggest that the distinction between acute and
long-lived SnCs warrants investigation in future experimental work. In
particular, future data collection efforts that capture spatial proximity
can help future modeling efforts better test the long-lived SnC SIS hypothesis.

More broadly, our success with applying a fully Bayesian approach to modeling
SnC growth trajectory illustrates the power of the Bayesian approach for
combining ODEs with robust uncertainty measurement. Relative to prior
work, our work provides more robust uncertainty estimates for biologically
relevant parameters and while enabling more principled model comparison.

# Conclusion
We developed a fully Bayesian model of SnC growth dynamics and fit it on
two datasets -- a longitudinal SnC growth trajectory dataset and a SnC
pulse dataset. We compared our model to Bayesian variants of Saturating
Removal and Unsaturated Removal models proposed by @karin2019senescent
and found that the SIS model fits the longitudinal dataset better than
both the SR and USR models but does not fit the pulse dataset as
well as the SR model, measured by both Bayes Factors and Leave-One-Out
Cross-Validation. We then estimated the median fraction of long-lived
SnCs in the longitudinal dataset and 5%-95% credible interval for that
fraction. Our results are preliminary but suggest that long-lived SnC
senescence-induced senescence is worth investigating as part of
future experimental work.

\pagebreak

# Appendix
## Differences between our and the original model {#model-diffs}
As mentioned in the main body, in order to reproduce @karin2019senescent's
model in the Bayesian framework and compare it to our SIS model, we had
to change the model in a few ways.

### Fully Bayesian vs. Maximum Likelihood model
The biggest change we made to the model, putting priors on parameters,
falls out of our adoption of the Bayesian paradigm. 

#### Priors
We put weakly informative half-normal priors on all ODE parameters with
prior hyperparameters either chosen based on stan's [Prior Choice Recommendations](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations)
or targeting a fraction of probability mass being within some range.

In all three models, \( \eta_0, \beta_0  \) we use \( \mathcal{N}^{+}(0, .78 \) priors to ensure
that 99% of the prior probability mass for initial growth / removal rates
is under 2. We chose 2 as the boundary because initial growth at this rate
would imply a turnover rate strongly at odds with domain knowledge and
the results of @karin2019senescent.

For \( \eta_1, \beta_1 \) we \( \mathcal{N}^{+}(0, 1) \) priors based on the
Prior Choice Recommendations document.

In the SIS model, for \( \alpha \), following analogous logic to what we
followed with \( \eta_0, \beta_0 \), we use a \( \rm{}(0, 0.025) \)
prior, which puts 99% of the probability mass on values below \( 0.02 \). This
is based on the assumption that long-lived SnCs shouldn't make up more than 60%
of SnCs at week 50. For \( l \), we use a \( \rm{HalfNormal}(0, .19) \)
prior under the now familiar assumption that 99% of our probability mass should
be on values below 99%. Again, this makes sense given that in young mice
it would be incredibly surprising for long-lived SnCs to make up more than
a relatively small fraction of all SnCs.

Finally, for our likelihood's standard deviation, we use a \( \mathsf{LogNormal}(-1, 1) \)
prior, following the multiplicative error model in @carpenter2018predator.

#### Approximate Posterior Uncertainty (vs. Analytical Approximations)
In @karin2019senescent, they use analytical approximation to determine
key uncertainty measures such as standard deviations of parameter estimates.
Here, we instead leverage posterior simulation to estimate parameter
uncertainty and generate retrodictions.

### Normally-distributed error with ODE vs. stochastic differential equation
The original model includes independent error terms for each time step,
with these error terms integrating to a Wiener process. Instead of time-specific
error terms, we use a normal likelihood parameterized by the output of
the integrated ODE at a given timestep and a single independent standard deviance
parameter (with a \( \rm{Log-Normal}(-1, 1) \) prior).
 
### Auto-catalysis term vs. SIS term
@karin2019senescent test a model that includes auto-catalysis, a different
name for what we call senescence-induced senescence. However, they found
that models that included auto-catalysis didn't fit the data as well as 
models that excluded. So why do our results differ from theirs? Although
it's hard to be certain the degree to which this is a result of our use of
a Bayesian approach vs. a maximum likelihood one, we believe our approach of 
explicitly modeling auto-catalysis
as only being driven by a fraction of the SnC population also plays a role.
Concretely, we suspect that including a separate term for initial long-lived SnC population
enabled our model to better disentangle overall SnC growth and SIS-driven growth.

\pagebreak

# References
